Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with square cross sections

### 2011-06-14 Text/paper.tex
Added rectangular channels results.

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\begin{document}
\begin{frontmatter}
\title{Binary-liquid lattice Boltzmann simulation of microchannels with square cross sections}
\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{kuzmin@ualberta.ca}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{Schlumberger DBR Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional case of the Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is a continuation of our previous work.  To validate an
applicability
of the LBM to Bretheron phenomena in the moderate range of the capillary number $0.05\leq Ca \leq
6.0$,
we focus on the simulation of three-dimensional capillaries with square cross sections.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to capture a number of phenomena occurin in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the dependency
of the bubble radii on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

\section{Introduction}
The Taylor/Bretherton \cite{bretherton} flow deals with long gas bubbles moving through liquid in
narrow channels. Depending on the channel geometry it was found that the deposited film thickness
is a complicated function of the capillary number $Ca$:
\begin{equation}
\label{capillary:number:definition}
Ca=\frac{U_{\mathrm{bubble}} \mu_{\mathrm{liq}}}{\gamma},
\end{equation}
where $U_{\mathrm{bubble}}$ is the bubble velocity, $\mu_{\mathrm{liq}}$ is the
dynamic liquid viscosity and $\gamma$ is the interfacial tension between gas and liquid.

For example, the film thickness
is proportional to $Ca^{2/3}$ in the range of small capillary numbers for circular channels
\cite{bretherton,heil-bretherton}. 
The problem of predicting flow patterns and associated mass transfers for Bretheron-type flows
is of significant interest for chemical industry as it is widely used in chemical monolith
microreactors \cite{kreutzer-pressure-drop}. Intensive mass transfer can be achieved because of the
large interfacial area and small diffusion lengths \cite{cerro-bubble-train}. Heat transfer is
also increased in comparison with single phase flow \cite{fukugata-levelset}. While it
is possible to calculate the flow analytically for small capillary numbers and nearly zero Reynolds
number \cite{bretherton}, the
results cannot be extrapolated to the wide range of capillary and Reynolds numbers used in the
chemical industry.  Thus, the need of consistent numerical simulations arises.

The two-dimensional geometry (circular tubes, parallel plates) flows are studied extensively in
experimental works of \citet{aussillous-deposition, cerro-bubble-train} and numerical works of
\citet{giavedoni-numerical,heil-bretherton}. All abovementioned investigations found that the 
Bretherton
analysis is valid only for small capillary numbers $Ca\leq 0.003$ and deviates from actual
measurements for larger ones.
This is caused by a complex interplay of gravitational,
interfacial, inertial and viscous forces \cite{gupta-review}. Historically, 
\citet{bretherton} neglected inertia effects. \citet{giavedoni-numerical} suggested
that
inertia effects
are negligible for $Ca \leq 0.05$ and have moderate impact for $Ca>0.05$ in the range of Reynolds
number from $0$ to $70$ for two-dimensional flows. Later on, \citet{heil-bretherton}
extended
the results for the flow between plates up to $Re=300$. This author noted that while the
influence of inertia on the established film thickness is
insignificant
($7$ percent deviation from the film thickness measured at $Re=0$), the change in Reynolds number
significantly influences  the pressure distrubution and the flow field near the front bubble cap.
\citet{cerro-bubble-train} showed that the mass transfer is strongly affected by flow direction for
small capillary numbers in case of upward and downward flows. Thus, to fully describe the
propagation of the semi-infinite finger in liquid media one needs to take into account the
viscous, gravitational, surface, and inertia forces \cite{gupta-review}.

In comparison with the two-dimensional Bretherton flows (circular tubes, flow between plates), there
is a vast number of experimental
results available for three-dimensional flows in microchannels with square, triangular and
rectangular cross sections. For instance, \citet{cerro-bubble-train} performed a number of
experiments for a
bubble-train flow in capillaries of
circular and square cross sections for horizontal, upward and downward flows.
\citet{shikazono-square} obtained
experimental
results for the film thickness dependence on the
capillary number for ethanol/air and water/air mixtures and for square, circular and triangular
shaped mirochannels. They also presented the experimental correlation for  bubble
radii
based on the capillary number $Ca$ and the Weber number $We=Re\,Ca$, where the Reynolds number is
defined as:
\begin{equation}
Re=\frac{U_{\mathrm{bubble}} H_{\mathrm{eff}}}{\nu_{\mathrm{liq}}},
\end{equation}
where $H_{\mathrm{eff}}$ is the channel height and $\nu_{\mathrm{liq}}$ is the kinematic liquid
viscosity.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular
or square-shaped capillaries there is a transition from an asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape is attributed to the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{Figures/capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{Figures/inset.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels indicating transition between non-axisymmetric case to
axisymmetric
one is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

There are also a number of numerical works on three-dimensional flows. For instance, 
\citet{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes for different
slug and channel cross sections.
\citet{heil-threedim} performed three-dimensional simulations for circular-,
square- and rectangular-shaped capillaries. They indicated a transition of flow regime in a liquid
slug. Vortexes observed at lower capillary numbers disappear at the capillary number above a
certain threshold,  Fig. \ref{fig:streamlines:pattern}. For a square channel the critical
capillary number is $Ca=0.691$.  \citet{heil-threedim} also found an empirical
correlation which
makes it possible to put on the same curve the results of radii dependency on the capillary number
simulations of rectangular channels with different aspect ratios.
It was also found that for microchannels with a certain aspect ratio $\alpha=\frac{a}{b}\geq
2.04$, the interface does not become
axisymmetric for any capillary number. \citet{wang-non-circular} performed numerical
simulations using the Volume of Fluid (VOF) technique for
capillaries with non-circular (square, triangular) cross sections. They also investigated the
relative slug to bubble velocity for
a range of capillary numbers.

As it was mentioned by \citet{gupta-review}, the most common techniques to simulate the Bretherton
phenomena are the VOF method \cite{wang-non-circular}, the level-set method
\cite{fukugata-levelset} and the finite element methods \cite{kreutzer-taylor,heil-threedim}. It was
also indicated \cite{gupta-review} that the new techniques, which are still in the development
stage,
are the lattice Boltzmann method and phase field methods \cite{anderson-diffuse,gurtin-binary}.
The finite element method solves the Bretherton flow as a free-surface problem with a sharp
interface but without underlying details for the gas.  The lattice Boltzmann method is a continuous
interface method, and therefore provides more flexibility in simulations involving coalescence
and/or
droplet breakup, and thus arises as a promising alternative for simulation of gas finger propagation.

@@

The binary liquid free-energy LB model due to \citet{swift} we used
simulates two liquids with the assumption of uniform overall
density. As it was indicated by \citet{gupta-review} a flow in microchannels is a complex
interplay of viscous, gravitational, inertial and surface forces, and the classical Bretherton
problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, the parameters in our simulations were carefully chosen to
avoid  inertia
effects, see Section \ref{sec:numerical:benchmark}. Moreover, the numerical results of
\citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers. Thus, a major governing factor for microchannel flows is not the density ratio, but
the
viscosity ratio. Therefore the uniform density binary liquid model is suitable for this kind of
simulations.
% THIS HAS ALREADY BEEN STATED BEFORE
% The goal of this work is to do a feasibility study of the LBM
% binary-liquid model to simulate and correctly predict flow patterns for the Bretherton/Taylor
% problem.

% TODO: This sentence should probably be moved to a summary/conclusions section?
The work results are in good agreement with other simulations \cite{heil-threedim,wang-non-circular}.

@@

The paper is organized as follows.  First, we explain the simulation benchmark construction
mainly based on our previous work \cite{kuzmin-binary2d}. Then, the binary liquid lattice
Boltzmann model is outlined. The simulation results for the three-dimensional case are presented in
the results section. That covers the film thickness dependency on the capillary number, the bubble
to slug relative velocity and vortex profiles. The paper
is concluded with a summary of the main findings.
%The critical capillary numbers for changing velocity pattern depending on the eccentricity
%parameter $\alpha$ are presented in Table \ref{table:recirculation:data}.
%\begin{table}
%\begin{tabular}{c|c}
%$\alpha$& $Ca_{T}$\\
%\hline
%1& 0.691\\
%1.1 & 0.688\\
%1.3 & 0.666\\
%1.5 & 0.631\\
%\end{tabular}
%\caption{The results for the recirculation region.\label{table:recirculation:data}}
%\end{table}

@@

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which require a
simulataneous imposement of pressure boundary conditions and the body force. The pressure boundary
conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
horizontal direction only and the upward/downward flows with pressure difference imposed are not
simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

It was indicated \cite{kuzmin-binary2d} that the film thickness for two-dimensional simulations should be at least twice as large
as the interface thickness. The interface thickness in the continuous formulation of the
free-energy binary liquid formulation occupies $3-5$ nodes. Therefore, the film should be
resolved as at least $6-10$ nodes.

\citet{heil-threedim} specify that the diagonal bubble radius is different from the
axial bubble radius for $Ca\leq \widehat{Ca}\approx 0.03$. The radii are the same for larger values
of
$Ca$ with a value $R_{diag}=R_{axis}=0.49 H_{\mathrm{eff}}$. Therefore, taking the minimal
requirement for the film thickness to be
resolved, i.e. $6$ lattice nodes, one would obtain a grid size of $600\times 600 \times 9000 $ to
properly resolve the binary liquid flow. This size implies that relatively
large
computational resources are required to perform the simulations.

One can argue that a non-uniform grid could be used. However, the film thickness is a
function of the curvatures of the front bubble tip \cite{bretherton} and the grid would need to
be refined just at the interface between gas and liquid. This is complicated to achieve for
dynamic systems where a bubble moves in the streamwise direction.
To reduce the computational overhead, we decided to take a simpler approach and exploit the
inherent symmetry of the problem by simulating only a quarter of the channel (see Appendix
\ref{append:sym} for details).

% TODO: This paragraph should be revisited, somehow it reads a little weirdly..
It should be noted that the binary liquid model, we used, has a uniform density. A part of
discussion is
addressed in the introduction:  to describe the Bretherton problem one needs to take into
account gravitational, viscous, inertial, and surface tension forces. Among all the works, there
is no full 
agreement between results even in the limit of Reynolds number $0$.  Moreover, as it was indicated
by \citet{kreutzer-taylor} and by \citet{cerro-bubble-train} the pressure distribution, streamlines
and bubble shape are strongly affected by the length of the liquid slug and the length of the
bubble. In current simulations we minimize the effects of bubbles train and inertia to have a least
impact, as the largest Reynolds number is not bigger than $10$ and the lengths of a bubble and a
channel are chosen to be as to minimize mutual effects of bubbles motion on each other and for the
film thickness to stabilize. However, more thorough studies are needed in order to understand
precisely the abovementioned effects.

@@

The next section briefly explains the binary liquid lattice Boltzmann model.
The steady-state approach, the dependency of the radii on the moderate capillary number values
$0.05\leq Ca \leq 6.0$, the velocity pattern, and the relative to the liquid slug bubble velocity 
are discussed in Section
\ref{sec:results}.

\section{Binary liquid lattice Boltzmann model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another in certain
directions specified by a discrete velocity set.
The LBE is typically implemented as follows \cite{ginzburg-boundary-main}:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{propagation step}\\
&g_i^{*}(\bm{x},t)=\omega_{\phi} g_i^{eq}(\bm{x},t)-(1-\omega_{\phi})
g_i(\bm{x},t),&&\text{collision step}\\
&g_i(\bm{x}+\bm{c_i},t+1)=g_i^{*}(\bm{x},t),&&\text{propagation step},
\end{aligned}
\end{equation}
where $\{f,g\}_i$ are the probability distribution functions in the direction $\bm{c_i}$, $\omega$
is the
relaxation parameter, $\omega_{\phi}$ is the phase relaxation parameter, and $F_i$ is the external
force population which
mimics the force behavior.

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\usepackage{tikz}
\begin{document}
\begin{frontmatter}
\title{Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with square
cross sections}
\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{kuzmin@ualberta.ca}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{Schlumberger DBR Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is the continuation of our previous work with simulations of
three-dimensional channels with square crosssections in the range of the capillary number $0.05\leq
Ca \leq
6.0$.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to correctly capture a number of phenomena occuring in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the way bubble radii
depend on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

\section{Introduction}
The Taylor/Bretherton \cite{bretherton} flow deals with long gas bubbles moving through liquid in
narrow channels. Depending on the channel geometry it was found \cite{gupta-review} that the 
thickness of the liquid film between a bubble and channel walls
is a complicated function of the capillary number $Ca$:
\begin{equation}
\label{capillary:number:definition}
Ca=\frac{U_{\mathrm{bubble}} \mu_{\mathrm{liq}}}{\gamma},
\end{equation}
where $U_{\mathrm{bubble}}$ is the bubble velocity, $\mu_{\mathrm{liq}}$ is the
dynamic liquid viscosity and $\gamma$ is the interfacial tension between gas and liquid.

For example, the film thickness
is proportional to $Ca^{2/3}$ for small capillary numbers for circular channels
\cite{bretherton,heil-bretherton}. 
Predicting flow patterns and associated mass transfer for Bretheron-type flows
is of significant interest for chemical industry as it is widely used in chemical monolith
microreactors \cite{kreutzer-pressure-drop}. Intensive mass transfer can be achieved because of the
large interfacial area and small diffusion lengths \cite{cerro-bubble-train}. Heat transfer is
also enhanced in comparison with single phase flow \cite{fukugata-levelset}. While it
is possible to calculate the flow analytically for small capillary numbers and nearly zero Reynolds
number \cite{bretherton}, such
results cannot be extrapolated to the wide range of capillary and Reynolds numbers used in the
chemical industry.  Thus, the need of consistent numerical simulations arises.

Flows in two-dimensional geometries (circular tubes, parallel plates) have been studied extensively
in the 
experimental works of \citet{aussillous-deposition}, and \citet{cerro-bubble-train}, and the
numerical works of
\citet{giavedoni-numerical,heil-bretherton}. All abovementioned investigations found that the 
Bretherton
analysis is valid only for small capillary numbers $Ca\leq 0.003$ and deviates from actual
measurements for larger $Ca$.
This is caused by a complex interplay of gravitational,
interfacial, inertial and viscous forces \cite{gupta-review}. Historically, 
\citet{bretherton} neglected inertia effects. \citet{giavedoni-numerical} suggested
that
inertia effects
are negligible for $Ca \leq 0.05$ and have moderate impact for $Ca>0.05$ in the range of Reynolds
number from $0$ to $70$ for two-dimensional flows. Later on, \citet{heil-bretherton}
studied the flow between plates up to $Re=300$. This author noted that while the
influence of inertia on the established film thickness is
insignificant
($7$ percent deviation from the film thickness measured at $Re=0$), the change in Reynolds number
significantly influences  the pressure distrubution and the flow field near the front bubble cap.
\citet{cerro-bubble-train} showed that the mass transfer is strongly affected by flow direction for
small capillary numbers in case of upward and downward flows. Thus, to fully describe the
propagation of the semi-infinite finger in liquid media one needs to take into account the
viscous, gravitational, surface, and inertia forces \cite{gupta-review}.

In comparison with the two-dimensional Bretherton flows (circular tubes, flow between plates), there
is a vast number of experimental
results available for three-dimensional flows in microchannels with square, triangular and
rectangular cross sections. For instance, \citet{cerro-bubble-train} performed a number of
experiments for a
bubble-train flow in capillaries of square cross sections for horizontal, upward and downward flows.
\citet{shikazono-square} obtained
experimental
results for the film thickness dependence on the
capillary number for ethanol/air and water/air mixtures and for square, circular and triangular
shaped mirochannels. They also presented an experimental correlation for  bubble
radii
based on the capillary number $Ca$ and the Weber number $We=Re\,Ca$, where the Reynolds number is
defined as:
\begin{equation}
Re=\frac{U_{\mathrm{bubble}} H_{\mathrm{eff}}}{\nu_{\mathrm{liq}}},
\end{equation}
where $H_{\mathrm{eff}}$ is the channel height and $\nu_{\mathrm{liq}}$ is the kinematic liquid
viscosity.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{Figures/capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{Figures/inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{Figures/legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

There are also a number of numerical works on three-dimensional flows. For instance, 
\citet{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes for different
slug and channel cross sections.
\citet{heil-threedim} performed three-dimensional simulations for circular-,
square- and rectangular-shaped capillaries. They indicated a transition of flow regime in a liquid
slug. Vortexes observed at lower capillary numbers disappear as the capillary number goes beyond a
certain threshold,  Fig. \ref{fig:streamlines:pattern}. For a square channel the critical
capillary number is $Ca=0.691$.  \citet{heil-threedim} also found an empirical
correlation which
makes it possible to collapse results of radii dependency on the capillary number
simulations of rectangular channels with different aspect ratios on a single curve.
It was also found that for microchannels with a certain aspect ratio $\alpha=\frac{a}{b}\geq
2.04$, the interface does not become
circular for any capillary number. \citet{wang-non-circular} performed numerical
simulations using the Volume of Fluid (VOF) technique for
capillaries with non-circular (square, triangular) cross sections. They also investigated the
relative slug to bubble velocity for
a range of capillary numbers.

As was mentioned by \citet{gupta-review}, the most common techniques to simulate the Bretherton
phenomena are the VOF method \cite{wang-non-circular}, the level-set method
\cite{fukugata-levelset} and the finite element methods \cite{kreutzer-taylor,heil-threedim}. It was
also indicated \cite{gupta-review} that the new techniques, which are still in the development
stage,
are the lattice Boltzmann method and phase field methods \cite{anderson-diffuse,gurtin-binary}.
The finite element method solves the Bretherton flow as a free-surface problem with a sharp
interface but without gas.  The lattice Boltzmann method is a continuous
interface method, and therefore provides more flexibility in simulations involving coalescence
and/or
droplet breakup, and thus arises as a promising alternative for simulation of gas finger propagation.

@@

The binary liquid free-energy LB model due to \citet{swift} that we used
simulates two liquids with the assumption of uniform overall
density. %As it was indicated by \citet{gupta-review} a flow in microchannels is a complex
%interplay of viscous, gravitational, inertial and surface forces. 
The classical Bretherton
problem is stated for gas and liquid, which are of
significantly
different density and viscosity. However, the numerical results of
\citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers, $Re<70$. The parameters in our simulations were carefully chosen to
avoid  inertia
effects, see Section \ref{sec:numerical:benchmark}. The maximum attained
Reynolds number $Re<10$. Thus, inertia effects can be neglected. Therefore, a major governing
factor for microchannel flows is not the density ratio, but
the
viscosity ratio. As a result, the uniform density binary liquid model is suitable for this kind of
simulations.  
% THIS HAS ALREADY BEEN STATED BEFORE
% The goal of this work is to do a feasibility study of the LBM
% binary-liquid model to simulate and correctly predict flow patterns for the Bretherton/Taylor
% problem.

% TODO: This sentence should probably be moved to a summary/conclusions section?
%The work results are in good agreement with other simulations
%\cite{heil-threedim,wang-non-circular}.

@@

The paper is organized as follows.  First, we explain the simulation benchmark construction
mainly based on our previous work \cite{kuzmin-binary2d}. Then, the binary liquid lattice
Boltzmann model is outlined. The simulation results for the three-dimensional case are presented in
the results section. That covers the film thickness dependency on the capillary number, the bubble
to slug relative velocity and vortex profiles for square microchannels. We also briefly cover flow
simulations for rectangular channels. The paper
is concluded with a summary of the main findings.
%The critical capillary numbers for changing velocity pattern depending on the eccentricity
%parameter $\alpha$ are presented in Table \ref{table:recirculation:data}.
%\begin{table}
%\begin{tabular}{c|c}
%$\alpha$& $Ca_{T}$\\
%\hline
%1& 0.691\\
%1.1 & 0.688\\
%1.3 & 0.666\\
%1.5 & 0.631\\
%\end{tabular}
%\caption{The results for the recirculation region.\label{table:recirculation:data}}
%\end{table}

@@

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which would require a
simultaneous imposement of pressure boundary conditions and the body force. 
%The pressure boundary
%conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
%horizontal direction only and the upward/downward flows with pressure difference imposed are not
%simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

It was indicated \cite{kuzmin-binary2d} that the film thickness for two-dimensional simulations should be at least twice as large
as the interface thickness. The interface thickness occupies $3-5$ nodes due to the continuous
interface formulation of the binary liquid model. Therefore, the
film should be
resolved as at least $6-10$ nodes.

\citet{heil-threedim} and Fig. \ref{fig:heil:three:dim} specify that the diagonal bubble radius is
different from the
axial bubble radius for $Ca\leq \widehat{Ca}\approx 0.03$. The radii are approximately the same for
larger values
of
$Ca$ with a value $R_{diag}=R_{axis}=0.49 H_{\mathrm{eff}}$. Therefore, taking the minimal
requirement for the film thickness to be
resolved, i.e. $6$ lattice nodes, one would obtain a grid size of $600\times 600 \times 9000 $ to
properly resolve the binary liquid flow. This size implies that relatively
large
computational resources are required to perform the simulations.

One can argue that a non-uniform grid could be used. However, the film thickness is a
function of the curvatures of the front bubble tip \cite{bretherton} and the grid would need to
be refined just at the interface between gas and liquid. This is complicated to achieve for
dynamic systems where a bubble moves in the streamwise direction.
To reduce the computational overhead, we decided to take a simpler approach and exploit the
inherent symmetry of the problem by simulating only a quarter of the channel (see
\ref{append:sym} for details).

% TODO: This paragraph should be revisited, somehow it reads a little weirdly..
It should be noted that the binary liquid model we used has a uniform density. This issue was
partially addressed in the introduction:  to describe the Bretherton problem one needs to take into
account gravitational, viscous, inertial, and surface tension forces. Among all the works, there
is no full 
agreement between results even in the limit of Reynolds number zero. Moreover,
as was indicated
by \citet{kreutzer-taylor} and by \citet{cerro-bubble-train} the pressure distribution, streamlines
and bubble shape are strongly affected by the length of the liquid slug and the length of the
bubble. In the current simulations we minimize the effects of bubbles train and inertia. Inertia
effect are minimized as the largest Reynolds number is not bigger than $10$. The mutual
effects of bubbles motion on each other is minimized by choosing lengths of a bubble and a
channel large enough in comparision with the channel height. However, more thorough studies are
needed in order to understand
precisely the abovementioned effects.

@@

%The next section briefly explains the binary liquid lattice Boltzmann model.
%The steady-state approach, the dependency of the radii on the moderate capillary number values
%$0.05\leq Ca \leq 6.0$, the velocity pattern, and the relative to the liquid slug bubble velocity 
%are discussed in Section
%\ref{sec:results}.

\section{Binary liquid lattice Boltzmann model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another in certain
directions specified by a discrete velocity set.
The LBE is typically implemented as follows \cite{ginzburg-boundary-main}:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{propagation step}\\
&g_i^{*}(\bm{x},t)=\omega_{\phi} g_i^{eq}(\bm{x},t)-(1-\omega_{\phi})
g_i(\bm{x},t),&&\text{collision step}\\
&g_i(\bm{x}+\bm{c_i},t+1)=g_i^{*}(\bm{x},t),&&\text{propagation step},
\end{aligned}
\end{equation}
where $\{f,g\}_i$ are the probability distribution functions in the direction $\bm{c_i}$, $\omega$
is the
relaxation parameter, $\omega_{\phi}$ is the phase relaxation parameter, and $F_i$ is the external
force population responsible for force inclusion to the Navier-Stokes equation.
